Applications of Many-Core Technologies to On-line Event Reconstruction in High Energy Physics Experiments

### 2013-11-20 papers/2013/NSS/NSS.tex
corrected a few typos

% disable the running header on all pages following the title page
\pagestyle{empty}
%
\begin{document}
% paper title
\title{Applications of Many-Core Technologies to On-line Event Reconstruction in High Energy Physics
Experiments}
% Author list
\author{A.~Gianelle, 
  S.~Amerio, 
  D.~Bastieri, 
  M.~Corvo, 
  W.~Ketchum,
  T.~Liu, 
  A.~Lonardo, 
  D.~Lucchesi,
  S.~Poprocki, 
  R.~Rivera, 
  L.~Tosoratto,
  P.~Vicini
  and 
  P.~Wittich
\thanks{Manuscript received November 15, 2013.
This work was supported by the  US National Science Foundation, the US 
Department of Energy Office of Science and the Italian
Istituto Nazionale di Fisica Nucleare. This work was partially supported by the 
EU Framework Programme 7 project EURETILE under grant number 247846.}% <-this % stops a space
\thanks{D.~Bastieri and D.~Lucchesi are with University of Padova and INFN Padova.}% 
\thanks{A.~Gianelle is with INFN Padova.}% 
\thanks{M.~Corvo is with University of Ferrara.}% 
\thanks{S.~Amerio is with University of Padova}% 
\thanks{W.~Ketchum is with Los Alamos National Laboratory.}%
\thanks{T.~Liu and R.~Rivera are with Fermi National Accelerator Laboratory.}%
\thanks{A.~Lonardo, L.~Tosoratto and P.~Vicini are with INFN Roma.}% 
\thanks{S.~Poprocki and P.~Wittich are with Cornell University.}
}
% make the title area
\maketitle
% after \maketitle, add the following command to remove the header on the title page
\thispagestyle{empty}
% abstract usually in passive voice
\begin{abstract}
  Interest in many-core architectures applied to real time selections is 
  growing in High Energy Physics (HEP) experiments. In this paper we describe
  performance measurements of many-core devices 
  when applied to a typical HEP online task: 
  the selection of events based on the trajectories of charged particles.
  We use as benchmark a scaled-up version of the algorithm used at CDF 
  experiment at Tevatron for online track reconstruction - the SVT algorithm - 
  as a realistic  test-case for low-latency trigger systems using new computing
  architectures for LHC experiment. We examine the
  complexity/performance trade-off in porting existing serial
  algorithms to many-core devices. We measure performance of
different architectures (Intel Xeon Phi and AMD GPUs, in addition 
to NVidia GPUs) and different software
environments (OpenCL, in addition to NVidia CUDA).
Measurements of both data processing  and data transfer latency are shown, 
considering different I/O strategies 
  to/from the many-core devices. \end{abstract}

@@

We showed initial studies on GPU performance in low-latency
environments ($\approx 100~\mu$s) in previous
papers~\cite{TIPP2011,NSS2012,CHEP2013}.  In this paper we extend
those studies to include other many-core architectures (Intel MIC in
addition to GPUs) and other programming toolsets (\textsc{opencl}).
The algorithm run on the parallel architecture is a complete version
of the fast track-fitting algorithm of the Silicon Vertex Tracker
(SVT) system at CDF~\cite{SVT1}.  We also investigate the performance
of the algorithm on a GPU manufactured by AMD using a different
computing environment: \textsc{opencl}.  Starting with a serial
algorithm implemented on a CPU, we test an \textit{embarrassingly
  parallel} algorithm on the Intel MIC environment. In this case each
event is handled independently by a core on the accelerator, and the
parallelization is achieved with only minor changes to the legacy
code.  This approach is only possible in the Intel MIC
environment. Next we consider an algorithm where we unroll three
internal nested loops and run these in parallel on a GPU, using the
\textsc{cuda} and \textsc{opencl} environments. This second approach
is programmatically more complicated and less trivial to implement. In
neither case have we re-thought the basic algorithms or the data
structures used. To achieve optimal performance, these steps would
have to be taken.  As one might expect, the improvement from the first
approach is rather modest, albeit easier to implement, and the second
approach shows larger performance gains. For GPUs, we also test
different strategies to transfer data to and from the device.

@@

To be run on NVidia GPUs, the code has been ported to \textsc{cuda}:
each step -- unpack, combine and track fit -- is performed by a
specific kernel; the three nested loops are unrolled so that each GPU
thread processes a single combination of hits.  The \textsc{cuda}
implementation makes use of \textsc{thrust}~\cite{bib_thrust}, a C++
template library for \textsc{cuda}, in the unpacking step. The
existence of template libraries such as \textsc{thrust} is an
advantage of the \textsc{cuda} environment.

To implement the algorithm to run on an AMD GPU, we have ported the
\textit{combine} and \textit{track fit} \textsc{cuda} kernels to
\textsc{opencl}, which requires minimal changes. Because the
\textsc{thrust} template libraries can only be used with
\textsc{cuda}, we resort to unpacking serially on the CPU.

@@

\section{Experimental setup and data flow}
The many-core devices used in this study are listed in
Table~\ref{tab_hwspecs}. The GPUs include a less expensive gaming
class GPUs (the NVidia GTX and AMD Radeon cards) and ones optimized
for scientific computing (Tesla). The MIC corresponds to a Xeon Phi
introduced in November 2012.

\begin{table*}[tbp]
  \caption{Capabilities of the many-core devices used in this study,
    according to the manufacturer's specifications. The first three
    are \textsc{nvidia} GPUs, the MIC 5110P is an Intel Xeon Phi, and
    the final one is an AMD GPU.  For Xeon Phi, the ``cores'' column
    counts the HW threads per core as equivalent to a GPU core.}
  \label{tab_hwspecs}
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
    Model & Tesla M2050 & Tesla K20m & GeForce GTX  Titan & MIC 5110P & Radeon HD 7970 \\
    \hline
     \hline
    Performance (SP, GFlops) & 1030 & 3520 & 4500 & 2022 & 3790\\
    Memory bandwidth  (GB/s) & 148 & 208 & 288  & 320 & 264\\   
    Memory size (GB) & 3 & 5 & 6 & 8 & 3\\
    Number of cores & 448 & 2496 & 2688 & 240 & 2048\\
    Clock speed (GHz) & 1.15 & 0.706 & 0.837 & 1.053 & 1.375\\
    \hline
  \end{tabular}
\end{table*}

@@

We have an additional setup for testing the \textsc{opencl}
implementation of the track fitting algorithm. Here, we use a 3.07 GHz
Intel Core i7 CPU 950, which has four cores and up to eight
computation threads. We run the algorithm serially on this CPU (one
core), and also run a CPU-based \textsc{opencl} algorithm which makes
use of the multi-core architecture. Additionally, we have an AMD
Radeon HD 7970 GPU in a PCIe slot in this setup, on which we also run
the \textsc{opencl} algorithm.

@@

\subsection{Data processing in \textsc{opencl}}
We also measure the track fitting algorithm latency in an
implementation using \textsc{opencl}. The \textsc{opencl} tools have
the advantage of being an open standard, while the \textsc{cuda} tools
are only compatible with Nvidia GPUs. Additionally, \textsc{opencl}
can work also on CPUs and on the Xeon Phi. For our tests, due to
limitations on the available space for storing data on the GPU, we
only test the algorithm on up to 1000 events (\textit{i.e.} about 2
million combinations to fit). In Fig.~\ref{fig:opencl_timing} we show
the results of algorithm latency measurements for three different
modes: running the serial algorithm on the CPU (single-core), running
the \textsc{opencl} algorithm on the CPU (multi-core), and running the
\textsc{opencl} algorithm on the AMD Radeon GPU. The \textsc{opencl}
implementation of the algorithm on the CPU provides a significant
speedup---about a factor of 5---over running the algorithm serially on
the CPU. Running the same \textsc{opencl} algorithm on the GPU
provides an even greater speedup due to the increased number of cores
and parallel threads that can be run, though there is additional
latency to copy data into and out of the GPU that makes running on the
GPU take longer for small numbers of combinations to fit.
\begin{figure}[!t]
  \centering
  \includegraphics[width=0.9\linewidth]{figures/TotalTimeOpencl_1000}
  \caption{The timing of the \textsc{opencl} algorithm as a function of the
    number of track fits. We compare timing running the algorithm serially on a CPU, 
    using \textsc{opencl} on the CPU, and on an AMD GPU.}
  \label{fig:opencl_timing}
\end{figure}
Surprisingly, though \textsc{opencl} is an open standard, we find that
we are unable to run the same \textsc{opencl} code on Nvidia GPUs. On two
different installations with two different video cards, we find
incorrect results running the code that ran successfully on the AMD
card. At present it is unclear what the root cause  is.

>>>>

% disable the running header on all pages following the title page
\pagestyle{empty}
%
\begin{document}
% paper title
\title{Applications of Many-Core Technologies to On-line Event Reconstruction in High Energy Physics
Experiments}
% Author list
\author{A.~Gianelle, 
  S.~Amerio, 
  D.~Bastieri, 
  M.~Corvo, 
  W.~Ketchum,
  T.~Liu, 
  A.~Lonardo, 
  D.~Lucchesi,
  S.~Poprocki, 
  R.~Rivera, 
  L.~Tosoratto,
  P.~Vicini
  and 
  P.~Wittich
\thanks{Manuscript received November 22, 2013.
This work was supported by the  US National Science Foundation, the US 
Department of Energy Office of Science and the Italian
Istituto Nazionale di Fisica Nucleare. This work was partially supported by the 
EU Framework Programme 7 project EURETILE under grant number 247846.}% <-this % stops a space
\thanks{D.~Bastieri and D.~Lucchesi are with University of Padova and INFN Padova.}% 
\thanks{A.~Gianelle is with INFN Padova.}% 
\thanks{M.~Corvo is with University of Ferrara.}% 
\thanks{S.~Amerio is with University of Padova.}% 
\thanks{W.~Ketchum is with Los Alamos National Laboratory.}%
\thanks{T.~Liu and R.~Rivera are with Fermi National Accelerator Laboratory.}%
\thanks{A.~Lonardo, L.~Tosoratto and P.~Vicini are with INFN Roma.}% 
\thanks{S.~Poprocki and P.~Wittich are with Cornell University.}
}
% make the title area
\maketitle
% after \maketitle, add the following command to remove the header on the title page
\thispagestyle{empty}
% abstract usually in passive voice
\begin{abstract}
  Interest in many-core architectures applied to real time selections is 
  growing in High Energy Physics (HEP) experiments. In this paper we describe
  performance measurements of many-core devices 
  when applied to a typical HEP online task: 
  the selection of events based on the trajectories of charged particles.
  We use as benchmark a scaled-up version of the algorithm used at CDF 
  experiment at Tevatron for online track reconstruction - the SVT algorithm - 
  as a realistic  test-case for low-latency trigger systems using new computing
  architectures for LHC experiment. We examine the
  complexity/performance trade-off in porting existing serial
  algorithms to many-core devices. We measure performance of
different architectures (Intel Xeon Phi and \textsc{amd} GPUs, in addition 
to NVidia GPUs) and different software
environments (OpenCL, in addition to NVidia CUDA).
Measurements of both data processing  and data transfer latency are shown, 
considering different I/O strategies 
  to/from the many-core devices. \end{abstract}

@@

We showed initial studies on GPU performance in low-latency
environments ($\approx 100~\mu$s) in previous
papers~\cite{TIPP2011,NSS2012,CHEP2013}.  In this paper we extend
those studies to include other many-core architectures (Intel MIC 
and \textsc{amd} GPUs in
addition to \textsc{nvidia} GPUs) and other programming 
toolsets (\textsc{opencl} in addition to \textsc{cuda}).
The algorithm run on the parallel architecture is a complete version
of the fast track-fitting algorithm of the Silicon Vertex Tracker
(SVT) system at CDF~\cite{SVT1}.  
%We also investigate the performance
%of the algorithm on a GPU manufactured by AMD using a different
%computing environment: \textsc{opencl}.  
Starting with a serial algorithm implemented on a CPU, 
we test an \textit{embarrassingly
  parallel} algorithm on the Intel MIC environment. In this case each
event is handled independently by a core on the accelerator, and the
parallelization is achieved with only minor changes to the legacy
code.  This approach is only possible in the Intel MIC
environment. Next we consider an algorithm where we unroll three
internal nested loops and run these in parallel on a GPU, using the
\textsc{cuda} and \textsc{opencl} environments. This second approach
is programmatically more complicated and less trivial to implement. In
neither case have we re-thought the basic algorithms or the data
structures used. To achieve optimal performance, these steps would
have to be taken.  As one might expect, the improvement from the first
approach is rather modest, albeit easier to implement, and the second
approach shows larger performance gains. For GPUs, we also test
different strategies to transfer data to and from the device.

@@

To be run on \textsc{nvidia} GPUs, the code has been ported to \textsc{cuda}:
each step -- unpack, combine and track fit -- is performed by a
specific kernel; the three nested loops are unrolled so that each GPU
thread processes a single combination of hits.  The \textsc{cuda}
implementation makes use of \textsc{thrust}~\cite{bib_thrust}, a C++
template library for \textsc{cuda}, in the unpacking step. The
existence of template libraries such as \textsc{thrust} is an
advantage of the \textsc{cuda} environment.

To implement the algorithm to run on an \textsc{amd} GPU, we have ported the
\textit{combine} and \textit{track fit} \textsc{cuda} kernels to
\textsc{opencl}, which requires minimal changes. Because the
\textsc{thrust} template libraries can only be used with
\textsc{cuda}, we resort to unpacking serially on the CPU.

@@

\section{Experimental setup and data flow}
The many-core devices used in this study are listed in
Table~\ref{tab_hwspecs}. The GPUs include a less expensive gaming
class GPUs (the \textsc{nvidia} GTX and \textsc{amd} Radeon cards) and ones optimized
for scientific computing (Tesla). The MIC corresponds to a Xeon Phi
introduced in November 2012.

\begin{table*}[tbp]
  \caption{Capabilities of the many-core devices used in this study,
    according to the manufacturer's specifications. The first three
    are \textsc{nvidia} GPUs, the MIC 5110P is an Intel Xeon Phi, and
    the final one is an \textsc{amd} GPU.  For Xeon Phi, the ``cores'' column
    counts the HW threads per core as equivalent to a GPU core.}
  \label{tab_hwspecs}
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
    Model & Tesla M2050 & Tesla K20m & GeForce GTX  Titan & MIC 5110P & Radeon HD 7970 \\
    \hline
     \hline
    Performance (SP, GFlops) & 1030 & 3520 & 4500 & 2022 & 3790\\
    Memory bandwidth  (GB/s) & 148 & 208 & 288  & 320 & 264\\   
    Memory size (GB) & 3 & 5 & 6 & 8 & 3\\
    Number of cores & 448 & 2496 & 2688 & 240 & 2048\\
    Clock speed (GHz) & 1.15 & 0.706 & 0.837 & 1.053 & 1.375\\
    \hline
  \end{tabular}
\end{table*}

@@

We have an additional setup for testing the \textsc{opencl}
implementation of the track fitting algorithm. Here, we use a 3.07 GHz
Intel Core i7 CPU 950, which has four cores and up to eight
computation threads. We run the algorithm serially on this CPU (one
core), and also run a CPU-based \textsc{opencl} algorithm which makes
use of the multi-core architecture. Additionally, we have an \textsc{amd}
Radeon HD 7970 GPU in a PCIe slot in this setup, on which we also run
the \textsc{opencl} algorithm.

@@

\subsection{Data processing in \textsc{opencl}}
We also measure the track fitting algorithm latency in an
implementation using \textsc{opencl}. The \textsc{opencl} tools have
the advantage of being an open standard, while the \textsc{cuda} tools
are only compatible with \textsc{nvidia} GPUs. Additionally, \textsc{opencl}
can work also on CPUs and on the Xeon Phi. For our tests, due to
limitations on the available space for storing data on the GPU, we
only test the algorithm on up to 1000 events (\textit{i.e.} about 2
million combinations to fit). In Fig.~\ref{fig:opencl_timing} we show
the results of algorithm latency measurements for three different
modes: running the serial algorithm on the CPU (single-core), running
the \textsc{opencl} algorithm on the CPU (multi-core), and running the
\textsc{opencl} algorithm on the \textsc{amd} Radeon GPU. The \textsc{opencl}
implementation of the algorithm on the CPU provides a significant
speedup---about a factor of 5---over running the algorithm serially on
the CPU. Running the same \textsc{opencl} algorithm on the GPU
provides an even greater speedup due to the increased number of cores
and parallel threads that can be run, though there is additional
latency to copy data into and out of the GPU that makes running on the
GPU take longer for small numbers of combinations to fit.
\begin{figure}[!t]
  \centering
  \includegraphics[width=0.9\linewidth]{figures/TotalTimeOpencl_1000}
  \caption{The timing of the \textsc{opencl} algorithm as a function of the
    number of track fits. We compare timing running the algorithm serially on a CPU, 
    using \textsc{opencl} on the CPU, and on an \textsc{amd} GPU.}
  \label{fig:opencl_timing}
\end{figure}
Surprisingly, though \textsc{opencl} is an open standard, we find that
we are unable to run the same \textsc{opencl} code on \textsc{nvidia} GPUs. On two
different installations with two different video cards, we find
incorrect results running the code that ran successfully on the \textsc{amd}
card. At present it is unclear what the root cause  is.
